---
title: "R Notebook"
output: html_notebook
---

You could use the `lme4` package, specifically the `lmer()` function to build a hierarchical model. 

## 1. Load Required Libraries

```{r}
library(lme4)
#library(lmerTest)  # Optional, for p-values in linear mixed models
```


### 2. Prepare Your Data

Let's consider:

* A dataset that includes variables `life_satisfaction`, `happy`, `age`, `gender`, `education`, and `country`.
* All Level 1 variables (individual level) are group-mean centered.
* `country` is a factor variable representing Level 2 units.

```{r}
# Group-mean centering Level 1 variables (optional but done in the ESS analysis)
data <- data %>%
  group_by(cntry) %>%
  mutate(
    happy = happy - mean(happy, na.rm = TRUE),
    age_c = agea - mean(agea, na.rm = TRUE),
    edu_c = eduyrs - mean(eduyrs, na.rm = TRUE)
  ) %>%
  ungroup()
```

### 3. Fit the HLM Using `lmer()`

```{r}
# Fit the multilevel model
model <- lmer(
  stflife ~ happy + age_c + gndr + edu_c + (1 | cntry),
  data = data
)
```

* `stflife` = Life satisfaction (DV, 0–10)
* `gndr` should be coded as a factor (e.g., Male/Female or 0/1)
* `(1 | cntry)` = Random intercepts for country

To view model output:

```{r}
summary(model)
```


### 4. Variance Explained


To understand how much of the variance lies between countries:

```{r}
null_model <- lmer(stflife ~ 1 + (1 | cntry), data = data)
VarCorr(null_model)
```

Calculate Intraclass Correlation Coefficient ICC:

  - quantifies the proportion of variance explained by the country 

  - The ICC is calculated by dividing the random effect variance by the total variance, i.e. the sum of the random effect variance and the residual variance
  
```{r}
icc <- as.numeric(VarCorr(null_model)$cntry[1]) / (as.numeric(VarCorr(null_model)$cntry[1]) + attr(VarCorr(null_model), "sc")^2)
icc
```


### Interpretation 

| Predictor       | R Code Equivalent | Interpretation                                                       |
| --------------- | ----------------- | -------------------------------------------------------------------- |
| Positive Affect | `happy`    | Strong positive effect on life satisfaction                          |
| Gender          | `gndr` (factor)   | Females (or whichever level is higher) have higher life satisfaction |
| Age             | `age_c`           | Not significant here ?                                                |
| Education       | `edu_c`           | Not significant here ?                                                |


